%%% Seznam použité literatury je zpracován podle platných standardů. Povinnou citační
%%% normou pro bakalářskou práci je ISO 690. Jména časopisů lze uvádět zkráceně, ale jen
%%% v kodifikované podobě. Všechny použité zdroje a prameny musí být řádně citovány.

\def\bibname{Bibliography}
\begin{thebibliography}{99}
\addcontentsline{toc}{chapter}{\bibname}

% Chapter 1

\bibitem{asn-basic}
	{\sc International Telecommunication Union}.
	ITU-T Recommendation X.680:
	\emph{Information technology -- Abstract Syntax Notation One (ASN.1): Specification of basic notation}.
	Geneva, Switzerland: ITU-T, 2008.
	ITU-T Recommendation series X, X.680.
	Available from: \url{http://www.itu.int/rec/T-REC-X.680-200811-I}

\bibitem{asn-vuln}
	CVE - Search Results: ASN.1.
	{\sc MITRE Corporation}.
	CVE - Common Vulnerabilities and Exposures
	[online].
	2013 [viewed 4 December 2013].
	Available from: \url{http://cve.mitre.org/cgi-bin/cvekey.cgi?keyword=ASN.1}

\bibitem{rfc5246}
	{\sc Dierks,} T. and {\sc Rescorla}, E.
	\emph{The Transport Layer Security (TLS) Protocol Version 1.2}
	[online].
	RFC 5246.
	Internet Society, 2008 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc5246}

% Chapter 2

\bibitem{rfc454}
	{\sc McKenzie,} A.
	\emph{File Transfer Protocol -- meeting announcement and a new proposed document}
	[online].
	RFC 454.
	Internet Society, 1973 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc454}

\bibitem{rfc959}
	{\sc Postel,} J. and {\sc Reynolds}, J.
	\emph{File Transfer Protocol}
	[online].
	RFC 959.
	Internet Society, 1985 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc959}

\bibitem{rfc4217}
	{\sc Ford-Hutchinson,} P.
	\emph{Securing FTP with TLS}
	[online].
	RFC 4217.
	Internet Society, 2005 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc4217}

\bibitem{rfc2616}
	{\sc Fielding,} R. et al.
	\emph{Hypertext Transfer Protocol -- HTTP/1.1}
	[online].
	RFC 2616.
	Internet Society, 1999 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc2616}

\bibitem{rfc2069}
	{\sc Franks,} J. et al.
	\emph{An Extension to HTTP : Digest Access Authentication}
	[online].
	RFC 2069.
	Internet Society, 1997 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc2069}

\bibitem{rfc2818}
	{\sc Rescorla,} E.
	\emph{HTTP Over TLS}
	[online].
	RFC 2818.
	Internet Society, 2000 [viewed 4 December 2013].
	Available from: \newline \url{http://tools.ietf.org/html/rfc2818}

\bibitem{rfc2518}
	{\sc Goland,} Y. et al.
	\emph{HTTP Extensions for Distributed Authoring -- WEBDAV}
	[online].
	RFC 2518.
	Internet Society 1999 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc2518}

\bibitem{rfc4918}
	{\sc Internet Engineering Task Force}.
	RFC 4918:
	\emph{HTTP Extensions for Web Distributed Authoring and Versioning (WebDAV)}
	[online].
	Edited by L. Dusseault.
	June 2007 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc4918}

\bibitem{rfc3253}
	{\sc Clemm,} G. et al.
	\emph{Versioning Extensions to WebDAV (Web Distributed Authoring and Versioning)}
	[online].
	RFC 3253.
	Internet Society, 2002 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc3253}

\bibitem{mssmb2}
	{\sc Microsoft Corporation}.
	[MS-SMB2]:
	\emph{Server Message Block (SMB) Protocol Versions 2 and 3}
	[online].
	Revision 44.0.
	Microsoft Corporation, 25 October 2013 [viewed 4 December 2013].
	Available from: \url{http://msdn.microsoft.com/en-us/library/cc246482.aspx}

\bibitem{nfs85}
	{\sc Sandberg,} Russel et al.
	\emph {Design and Implementation of the Sun Network Filesystem}.
	USENIX, 1985.
	Available from: \newline \url{http://pdos.csail.mit.edu/6.824-2004/papers/sandberg-nfs.pdf}

\bibitem{rfc1094}
	{\sc Nowicki,} B.
	\emph{NFS: Network File System Protocol specification}
	[online].
	RFC 1094.
	Internet Society, 1989 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc1094}

\bibitem{rfc1813}
	{\sc Callaghan,} B., {\sc Pawlowski,} B., and {\sc Staubach}, P.
	\emph{NFS Version 3 Protocol Specification}
	[online].
	RFC 1813.
	Internet Society, 1995 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc1813}

\bibitem{rfc1057}
	{\sc Sun Microsystems}.
	\emph{RPC: Remote Procedure Call Protocol specification: Version 2}
	[online].
	RFC 1057.
	Edited by .
	Internet Society, 1988 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc1057}

\bibitem{rfc1832}
	{\sc Srinivasan,} R.
	\emph{XDR: External Data Representation Standard}
	[online].
	RFC 1832.
	Internet Society, 1995 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc1832}

\bibitem{juszczak89}
	{\sc Juszczak,} Chet.
	Improving the Performance and Correctness of an NFS Server.
	In: {\emph Proceedings of the Winter 1989 USENIX Technical Conference}.
	USENIX, 1989, pp. 53-63.

\bibitem{rfc5661}
	{\sc Internet Engineering Task Force}.
	RFC 5661:
	\emph{Network File System (NFS) Version 4 Minor Version 1 Protocol}
	[online].
	Edited by S. Shepler, M. Eisler, and D. Noveck.
	January 2010 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc5661}

\bibitem{rfc2339}
	{\sc The Internet Society} and {\sc Sun Microsystems}.
	RFC 2339:
	\emph{An Agreement Between the Internet Society, the IETF, and Sun Microsystems, Inc. in the matter of NFS V.4 Protocols}
	[online].
	May 1998 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc2339}

\bibitem{rfc4251}
	{\sc Ylonen,} T.
	\emph{The Secure Shell (SSH) Protocol Architecture}
	[online].
	RFC 4251.
	Edited by C. Lonvick.
	January 2006 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc4251}

\bibitem{secsh-email}
	{\sc Hartman,} Sam.
	Formal consultation prior to closing the secsh working group.
	In: \emph{ietf.secsh e-mail conference}
	[online].
	30 July 2006; 19:15:10 -0400 [viewed 4 December 2013; 18:32 CET].
	Message-ID: tslvepgknpd.fsf@cz.mit.edu.
	Available from: \newline \url{http://osdir.com/ml/ietf.secsh/2006-07/msg00010.html}

\bibitem{draft-secsh-13}
	{\sc Galbraith,} J. and {\sc Saarenmaa,} O.
	\emph{SSH File Transfer Protocol, draft-ietf-secsh-filexfer-13.txt}
	[online].
	Internet-Draft.
	10 July 2006 [viewed 4 December 2013].
	Available from: \newline \url{http://tools.ietf.org/html/draft-ietf-secsh-filexfer-13}

% Chapter 3

\bibitem{rfc4422}
	{\sc Internet Engineering Task Force}.
	RFC 4422:
	\emph{Simple Authentication and Security Layer (SASL)}
	[online].
	Edited by A. Melnikov and K. Zeilenga.
	June 2006 [viewed 4 December 2013].
	Available from: \newline \url{http://tools.ietf.org/html/rfc4422}

\bibitem{kelsey2002}
	{\sc Kelsey,} John.
	Compression and Information Leakage of Plaintext.
	In: \emph{Fast Software Encryption, 9th International Workshop, FSE 2002,
	Leuven, Belgium, February 4-6, 2002, Revised Papers}
	Berlin: Springer Berlin Heidelberg, 2002, pp. 263--276.
	ISBN 978-3-540-45661-2.
	Available from: \url{http://www.iacr.org/cryptodb/archive/2002/FSE/3091/3091.pdf}

\bibitem{pep383}
	{\sc v. Löwis,} Martin.
	\emph{Non-decodable Bytes in System Character Interfaces}
	[online].
	PEP 383.
	Python Software Foundation, 2012 [viewed 4 December 2013].
	Available from: \url{http://www.python.org/dev/peps/pep-0383/}

\bibitem{rfc2743}
	{\sc Linn,} J.
	\emph{Generic Security Service Application Program Interface Version 2, Update 1}
	[online].
	RFC 2743.
	January 2000 [viewed 4 December 2013].
	Available from: \url{http://tools.ietf.org/html/rfc2743}

\bibitem{xwindow}
	{\sc Scheifler,} Robert W. and {\sc Gettys,} James.
	\emph{X Window System: Core and extension protocols: X version 11, releases 6 and 6.1}.
	Edited by Al Mento and Donna Converse.
	Boston: Digital Press, 1997.
	ISBN 978-1-55558-148-0

%\bibitem{rfcXX}
%	{\sc Internet Engineering Task Force}.
%	RFC XX:
%	\emph{}
%	[online].
%	Edited by .
%	 [viewed 4 December 2013].
%	Available from: \url{http://tools.ietf.org/html/rfcXX}


\end{thebibliography}
